Toward Automatic Translation from OpenACC to OpenMP 4

### 2016-04-24 sig-alternate-sample.tex
Added grammar and translation rules

@@

\todo{0.5 page - description of the conversion algorithm (what clauses translate to what?) - Nawrin/Dr.Overbey} 
\section{Evaluation}
\todo{1 page including tables - evaluation: what code did we test it on; how many lines changed; what performance numbers did we get - Nawrin}
\subsection{Test Corpus}
\subsection{Performance}

>>>>

\subsection{A Subset of OpenACC}
For the purposes of prototyping, we designed our tool to accept the subset of
OpenACC described by the grammar in Figure~\ref{fig:grammar}.  This includes
all of the directives and clauses used by the EPCC Benchmarks, as well as the
other example programs in our test suite.  While it omits some important
features---asynchronous execution, unstructured data lifetimes, etc.---it does
correspond to a useful, commonly-used subset of OpenACC.

\begin{figure}
\begin{tabular}{rcl}
\textit{acc-directive} &$\rightarrow$& \texttt{\#pragma acc data\ }\textit{data-clauses} \\
                       &     $|$     & \texttt{\#pragma acc parallel}\texttt{\ }\textit{parallel-clauses} \\
                       &     $|$     & \texttt{\#pragma acc loop\ }\textit{type}\texttt{\ }\textit{loop-clauses} \\
\textit{type}          &$\rightarrow$& \texttt{gang} \\
                       &     $|$     & \texttt{vector} \\
                       &     $|$     & \texttt{seq} \\
                       &     $|$     & $\epsilon$ \\
\textit{data-clause}   &$\rightarrow$& \texttt{copyin(}\textit{vars}\texttt{)} \\
                       &     $|$     & \texttt{copyout(}\textit{vars}\texttt{)} \\
                       &     $|$     & \texttt{copy(}\textit{vars}\texttt{)} \\
                       &     $|$     & \texttt{create(}\textit{vars}\texttt{)} \\
\textit{parallel-clause}&$\rightarrow$&\texttt{num\_gangs(}\textit{n}\texttt{)} \\
                       &     $|$     & \texttt{vector\_length(}\textit{n}\texttt{)} \\
                       &     $|$     & \texttt{private(}\textit{vars}\texttt{)} \\
                       &     $|$     & \texttt{reduction(}\textit{reductions}\texttt{)} \\
\textit{loop-clause}   &$\rightarrow$& \texttt{independent} \\
                       &     $|$     & \texttt{private(}\textit{vars}\texttt{)} \\
\end{tabular}
\caption{OpenACC subset grammar accepted by the translation tool.  Syntax is
omitted for \textit{vars}, \textit{reductions}, and integers \textit{n}.  The
nonterminals \textit{data-clauses} and \textit{loop-clauses} correspond to
sequences of zero or more phrases derived from \textit{data-clause} and
\textit{loop-clause}, respectively.  In addition to the directives shown,
\texttt{\#pragma acc parallel loop} is supported, and treated as syntactic
sugar for a \texttt{loop} directive nested directly under a \texttt{parallel}
directive.}
\label{fig:grammar}
\end{figure}

\subsection{Considerations in Translation}
In theory, the goals of the present paper are modest: We are interested in
translating a subset of OpenACC (Figure~\ref{fig:grammar}) to OpenMP~4,
targeting the same accelerator devices (NVIDIA GPUs).  In theory, the two APIs
are very closely aligned, and the translation should be straightforward.

In practice, the situation is not so simple.  The OpenACC and OpenMP
specifications define an \emph{abstract} execution model.  They do not dictate
how that model is mapped to specific hardware devices.  This is left to the
compiler.  Our OpenACC programs were compiled using PGCC~16.1, the standard
commercial compiler for OpenACC\@.  We compiled OpenMP~4 using a custom build
of Clang~3.8\footnote{See \url{https://parallel-computing.pro/index.php/9-cuda/43-openmp-4-0-on-nvidia-cuda-gpus}} (PGCC does not yet support
OpenMP~4\footnote{Cray's compilers support both, but there are currently no
Cray resources in the XSEDE ecosystem.}).  As we will show, directives that
appear similar may be translated quite differently by the two compilers.

\subsubsection{Work Distribution Differs}

\begin{figure*}
\begin{center}
\begin{minipage}{4in}
{\footnotesize
\begin{verbatim}
#pragma acc data copyin(A[0:N], B[0:N]), copyout(C[0:N])
#pragma omp target data map(to:A[0:N], B[0:N]), map(from:C[0:N])
{
    #pragma acc parallel loop
    #pragma omp target teams distribute parallel for
    for (int i = 0; i < N; ++i)
    {
        C[i] = A[i] + B[i];
    }
}
\end{verbatim}
}
\end{minipage}
\end{center}
\caption{Vector addition}
\label{fig:vecadd}
\end{figure*}

Perhaps the simplest example of directive-based accelerator code is the vector
addition kernel shown in Figure~\ref{fig:vecadd}.  It adds two vectors $A$ and
$B$ on an accelerator device, storing the sum in $C$.  While the directives
direct the compiler to parallelize the loop on the accelerator, the compiler
can decide (in this case) how to divide the loop iterations among threads
and thread blocks on the CUDA device.  When $N = 10000$:
\begin{itemize}
\item PGCC generates a grid of 79 thread blocks with 128 threads per block;
kernel execution time is about $3~\mu s$ on a Tesla~K40.
\item Clang generates a grid of 15 thread blocks with 1024 threads per block;
kernel execution time is about $30~\mu s$ on a Tesla~K40.\footnote{Adding
\texttt{num\_teams(79) thread\_limit(128)} to the OpenMP code (to match the
grid and block dimensions of the OpenACC code) actually increases the kernel
execution time to about $80~\mu s$.
%Adding \texttt{num\_gangs(15) vector\_length(1024)} to the OpenACC code does
%not change the performance significantly.
}
\end{itemize}

It is also worth noting that PGCC generates code where all threads are active
in all thread blocks except the last (presumably to minimize control
divergence).  Clang generates code that assigns an approximately equal number
of threads to each thread block, so inactive threads may be present in
several blocks.

\begin{figure*}
\begin{minipage}{.5\textwidth}
{\footnotesize
\begin{verbatim}
// OpenACC
#pragma acc parallel loop
for (int i = 0; i < 10; i++) {      // Gang
    for (int j = 0; j < 10; j++) {  // Vector
        ...
    }
}

\end{verbatim}
}
\end{minipage}
\begin{minipage}{.5\textwidth}
{\footnotesize
\begin{verbatim}
// OpenMP 4 Device Directives
#pragma omp target teams distribute
for (int i = 0; i < 10; i++) {
    #pragma omp parallel for
    for (int j = 0; j < 10; j++) {
        ...
    }
}
\end{verbatim}
}
\end{minipage}
\caption{Doubly nested loops.}
\label{fig:nest}
\end{figure*}

More important is the handling of loop nests, such as the simple
doubly nested loops in Figure~\ref{fig:nest}.

In OpenACC, the compiler has some freedom to decide how to parallelize the loop
nest.  PGCC will typically run the outermost loop as a gang loop and the
innermost loop as a vector loop.  Given the OpenACC code in
Figure~\ref{fig:nest}, the compiler does exactly this: the iterations of the
$i$-loop are divided among the thread blocks, and the iterations of the
$j$-loop are divided among the threads within each block.

In OpenMP, work distribution is not so automatic.  Consider these
``obvious'' but incorrect translations of the OpenACC code into OpenMP: The
\texttt{\#pragma acc parallel loop} directive could be replaced by
\begin{itemize}
\item \texttt{\#pragma omp target teams}.  In this case, the entire loop
nest would be run redundantly by all thread blocks.
\item \texttt{\#pragma omp target teams distribute}.  In this case, the
iterations of the $i$-loop would be divided among thread blocks, but only
one thread would be active within each block.
\item \texttt{\#pragma omp target teams distribute parallel for}.  Here,
the iterations of the $i$-loop are distributed among threads and thread blocks.
\end{itemize}
In all three cases, \emph{the inner $j$-loop runs sequentially}, because the
parallelization directive is applied to only one loop.  OpenMP requires the
programmer to explicitly specify how to run each loop in the nest.  The
translation most similar to the OpenACC code is shown on the right in
Figure~\ref{fig:nest}: The iterations of the $i$-loop are divided among thread
blocks using the \texttt{distribute} directive, and the iterations of the
$j$-loop are divided among threads in the block using the \texttt{parallel for}
directive.

%\subsubsection{Intrinsics Differ}

%\subsubsection{Reductions Differ}

\subsubsection{Data Transfer May Differ}

In both OpenACC and OpenMP, if \texttt{data}/\texttt{target data} directives
are omitted, the compiler will automatically determine what data needs to be
transferred between the host and the accelerator.  This is another case where
compilers may differ.  In the vector addition example
(Figure~\ref{fig:vecadd}), consider what happens when $A$, $B$, and $C$ are
\emph{pointers} allocated just before the kernel (in the same function) using
\textit{malloc}.  With PGCC, the \texttt{\#pragma acc data} directive can be
omitted; it can determine statically that 10,000 elements of $A$, $B$, and
$C$ are accessed and thus need to be copied.  With Clang, if the
\texttt{\#pragma omp target data} directive is omitted, the kernel crashes at
runtime (\emph{Device Illegal Address}).

\subsection{Translation Algorithm}

\newcommand{\translate}[1]{\ensuremath{\mathcal{T}[\![ #1 ]\!]}}
\begin{figure*}
\begin{align*}
%
\translate{\texttt{\#pragma acc data\ }\textit{data-clauses}} &=
   \texttt{\#pragma omp target data } \translate{\textit{data-clauses}} \\
%
\translate{\texttt{copyin(}\textit{vars}\texttt{)}} &=
   \texttt{map(to:}\textit{vars}\texttt{)} \\
%
\translate{\texttt{copyout(}\textit{vars}\texttt{)}} &=
   \texttt{map(from:}\textit{vars}\texttt{)} \\
%
\translate{\texttt{copy(}\textit{vars}\texttt{)}} &=
   \texttt{map(tofrom:}\textit{vars}\texttt{)} \\
%
\translate{\texttt{create(}\textit{vars}\texttt{)}} &=
   \texttt{map(alloc:}\textit{vars}\texttt{)} \\
%
\translate{\texttt{\#pragma acc parallel\ }\textit{parallel-clauses}} &=
   \left.\parbox[t]{4in}{%
   \texttt{\#pragma omp target teams\ }\translate{\textit{parallel-clauses}}}\right. \\
%
\translate{\texttt{num\_gangs(}\textit{n}\texttt{)}} &=
   \texttt{num\_teams(}\textit{n}\texttt{)} \\
%
\translate{\texttt{vector\_length(}\textit{n}\texttt{)}} &=
   \texttt{thread\_limit(}\textit{n}\texttt{)} \\
%
\translate{\texttt{private(}\textit{vars}\texttt{)}} &=
   \texttt{private(}\textit{vars}\texttt{)} \\
%
\translate{\texttt{reduction(}\textit{vars}\texttt{)}} &=
   \texttt{reduction(}\textit{reductions}\texttt{)} \\
%
\translate{\texttt{\#pragma acc loop\ }\textit{type}\texttt{\ }\textit{loop-clauses}} &=
   \begin{cases}
   \texttt{\#pragma omp distribute\ }\translate{\textit{loop-clauses}} \\
        \hspace*{2em} \textup{ if $\textit{type} \neq \texttt{seq}$, and the loop is the outermost} \\
        \hspace*{3em} \textup{ non-sequential loop in a loop nest} \\
        \hspace*{3em} \textup{ in a \texttt{parallel} region} \vspace*{.25em} \\
   \texttt{\#pragma omp parallel for\ }\translate{\textit{loop-clauses}} \\
        \hspace*{2em} \textup{ if $\textit{type} \neq \texttt{seq}$, and the loop is the innermost} \\
        \hspace*{3em} \textup{ non-sequential loop in a loop nest} \\
        \hspace*{3em} \textup{ in a \texttt{parallel} region} \vspace*{.25em} \\
   \texttt{\#pragma omp distribute parallel for\ }\translate{\textit{loop-clauses}} \\
        \hspace*{2em} \textup{ if both of the above apply} \vspace*{.33em} \\
   \epsilon
        \hspace*{1.65em} \textup{ otherwise} \\
   \end{cases} \\
%
\translate{\texttt{independent}} &=
   \epsilon \\
%
\end{align*}
\caption{Translation rules from OpenACC to OpenMP~4.  The function
$\mathcal{T}$ maps phrases derived from the OpenACC subset grammar to
strings corresponding to one or more OpenMP directives.  Sequences
(e.g., \textit{parallel-clauses}) are translated elementwise, that is,
$\translate{t_1\ t_2\ \dots\ t_n} =
\translate{t_1}\ \translate{t_2}\ \dots\ \translate{t_n}$.}
\label{fig:rules}
\end{figure*}

In the previous section, we showed that OpenACC and OpenMP~4 device directives
handle work distribution differently, and that differences in compilers can
have significant effects on the performance (and correctness) of code.  This
means that porting from OpenACC to OpenMP~4 will almost certainly involve
some manual effort.  Nevertheless, much of the translation is straightforward
and tedious, which makes it worth automating.

Our tool implements the translation rules in Figure~\ref{fig:rules}.  The
majority of the translation is straightforward: Each phrase in OpenACC is
mapped to a corresponding phrase in OpenMP\@.  These translation rules were
designed with the following objectives:
\begin{itemize}
\item \emph{The translation should be straightfoward and consistent.}
Each OpenACC directive should translate to the same OpenMP directive(s)
every time, and the translation should mirror the structure of the OpenACC
code, to the extent possible.
\item \emph{Automated translation and manual editing should be complementary.}
The tool should not try to automate too much.  It should not perform loop
nest optimizations; it should not infer implicit data transfers; it should
not make any invasive code changes.  The tool should automate the tedious,
most obvious parts of the translation, in a completely predictable way.
More sophisticated changes---like performance tuning---should be under the
control of the programmer.
\end{itemize}

\subsubsection{Translating Data Directives}

It is important to note that, according to the translation rules in
Figure~\ref{fig:rules}, \emph{each data directive is translated as-is}.  No
attempt is made to infer implicit data transfers or array lengths.

\subsubsection{Translating Loop Nests}

OpenACC's \texttt{gang} clause is roughly equivalent to OpenMP's
\texttt{distribute}---effectively, it distributes work among thread blocks.
Likewise, OpenACC's \texttt{vector} clause is roughly equivalent to OpenMP's
\texttt{parallel for} inside a \texttt{teams} region: it distributes work among
threads within a block.

Unfortunately, translation isn't quite so simple.  OpenACC does not
\emph{require} the \texttt{gang} and \texttt{vector} clauses to be used, in
which case the work distribution is determined by the compiler.  OpenACC is
also surprisingly lax about where the nesting of gang and vector loops.  For
example, a \texttt{gang} loop may be nested inside a \texttt{vector} loop;
there is no direct equivalent of this in OpenMP.  Moreover, PGCC may ignore the
\texttt{gang} and \texttt{vector} directives altogether, e.g., when a gang loop
appears inside another gang loop, or a vector loop appears outside another
vector loop.

For these reasons, \emph{the loop-handling rule of Figure~\ref{fig:rules}
ignores OpenACC's \texttt{gang} and \texttt{vector} clauses}.  Instead, it uses
a heuristic typical of parallelizing compilers (one that PGCC appears to use,
in fact): the iterations of the outermost parallelizable loop are partitioned
among thread blocks, and the iterations of the innermost parallelizable loop
are partitioned among the CUDA threads within that block.

@@

\todo{0.5 page - description of the conversion algorithm (what clauses translate to what?) - Nawrin/Dr.Overbey} 
\section{Evaluation}

PGCC~16.1 and Clang~3.8

\todo{1 page including tables - evaluation: what code did we test it on; how many lines changed; what performance numbers did we get - Nawrin}
\subsection{Test Corpus}
\subsection{Performance}
